---
title: "Advanced Phenotyping in Single Cell Data Analysis"
author: Yingxin Lim^[yinxin.lin@sydney.edu.au]
        Hani Jieun Kim^[hani.kim@sydney.edu.au]
        Carissa Chen^[cchen@uni.sydney.edu.au]
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An Example Workshop}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Advanced Phenotyping in Single Cell Data Analysis

Authors:
    Yingxin Lim^[The University of Sydney],
    Hani Jieun Kim^[The University of Sydney],
    Carissa Chen^[The University of Sydney].
    <br/>
Last modified: 22 Sept, 2021.

## Overview

### Description

[Brief description of workship]

### Pre-requisites

List any workshop prerequisites, for example:

* Basic knowledge of R syntax
* Familiarity with the GenomicRanges class
* Familiarity with xyz vignette (provide link)

List relevant background reading for the workshop, including any
theoretical background you expect students to have.

* List any textbooks, papers, or other reading that students should be
  familiar with. Include direct links where possible.

### Participation

Describe how students will be expected to participate in the workshop.

### _R_ / _Bioconductor_ packages used

List any _R_ / _Bioconductor_ packages that will be explicitly covered.

### Time outline

Structure of the 2-hour workshop:

| Activity                     | Time |
|------------------------------|------|
| Packages                     | 15m  |
| Package Development          | 15m  |
| Contributing to Bioconductor | 5m   |
| Best Practices               | 10m  |

### Workshop goals and objectives

List "big picture" student-centered workshop goals and learning
objectives. Learning goals and objectives are related, but not the
same thing. These goals and objectives will help some people to decide
whether to attend the conference for training purposes, so please make
these as precise and accurate as possible.

*Learning goals* are high-level descriptions of what
participants will learn and be able to do after the workshop is
over. *Learning objectives*, on the other hand, describe in very
specific and measurable terms specific skills or knowledge
attained. The [Bloom's Taxonomy](#bloom) may be a useful framework
for defining and describing your goals and objectives, although there
are others.

### Learning goals

Some examples:

* describe how to...
* identify methods for...
* understand the difference between...

### Learning objectives

* analyze xyz data to produce...
* create xyz plots
* evaluate xyz data for artifacts

## 1. Introduction

General introduction to the advanced phenotyping course (e.g., summary of the introduction PPT, Carissa to fill?).

## 2. Defining cellular identity

### 2.1 Load libraries
```{r}

```

### 2.2 Loading preprocessed data
About the data... from [this paper](https://www.nature.com/articles/s41586-018-0414-6#additional-information). We provide the preprocessed version in this workshop. The original data can be access from GEO with the accession number XXXXXX.
```{r}
# code to load data 
# we can also run some of the processing steps here (e.g., log transform, filter lowly expressed genes etc.)
```

### 2.3 Visualising the data
```{r}
# code to run PCA/UMAP/TSNE
```

### 2.3 Clustering the data
```{r}
# code to perform clustering of the data
```


### 2.3 Identifying the cell types

Now, we want to identify a cluster to a cell type. This task is easy to apply when discrete and well-annotated cell types, i.e., those with well-known markers, are present in your data. The procedure becomes more challenging when we are met with a data with rare or unknown cell types or transitional cell states. Either way, the general idea is to find a set of genes that are expressed different in a cell compared to the rest of the cells in the dataset. Thus, gene sets must be identifying of the particular cell type of the cell.

There are many methods

#### Cepo to identify cell identity genes
```{r}
# Code to run Cepo
```

#### Examine expression of Cepo genes in the low dimensional space
```{r}
# Code to visualise markers
```

#### Violin plots

Violin plots enable us to examine the distribution of the expression values associated with each cell in the clusters. The x-axis denote the clusters, and the y-axis represents the log-transformed gene expression of a given gene. Alternatives of violin plots are jitter and swarmplots, which unlike violin plots show each cell as a point.
```{r}
# Code to plot violin plot
```

#### Saving outputs
Finally, we can save our marker gene analysis for use later. These top marker genes can be used to subset the data prior to performing classification of the data
```{r}
# Code to save top X genes 
```

## 3. Defining a trajectory

## 4. Characterising a trajectory

## 5. Cell-cell communication

## 6. Conclusions and wrap-up




